Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression

### 2022-06-26 docs/rmd/paper.Rmd
autogit
...

```{r setup, include=FALSE}
library(kableExtra)
library(magrittr)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.pos = "!H")
```

@@

> $H0_{1}$: The distribution underlying the samples of training/testing errors produced by torunament selection is the same as the distribution underlying samples of training/testing errors produced by $\epsilon$-lexicase selection

@@

> $H0_{2}$: The distribution underlying the samples of training errors produced by $\epsilon$-lexicase/torunament selection is the same as the distribution underlying samples of testing errors produced by $\epsilon$-lexicase/tournament selection

>>>>

```{r setup, include=FALSE}
library(kableExtra)
library(magrittr)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.pos = "!H")
```
\setcounter{page}{1}

@@

> $H0_{1}$: The distribution underlying the samples of training/testing errors produced by tournament selection is the same as the distribution underlying samples of training/testing errors produced by $\epsilon$-lexicase selection

@@

> $H0_{2}$: The distribution underlying the samples of training errors produced by $\epsilon$-lexicase/tournament selection is the same as the distribution underlying samples of testing errors produced by $\epsilon$-lexicase/tournament selection

### 2022-06-26 docs/rmd/presentation/slides.Rmd
autogit
...

* Recent alternative: Lexicase Selection and it's variation $\epsilon$-lexicase selection
* Idea: Selection method for uncompromising, continous-valued symbolic regression problems ^[@6920034, p.12]
* Increases genetic diversity inside the population^[@6920034, p.1]
* Higher chance for "specialist" solutions to be selected since it is decided on a per case basis
* Performance increases have been demosntrated in many benchmarking problems ^[@epsilon_lexicase_main, p.744-745]

@@

>>>>

* Recent alternative: Lexicase Selection and it's variation $\epsilon$-lexicase selection
* Idea: Selection method for uncompromising, continous-valued symbolic regression problems ^[@6920034, p.12]
* Increases genetic diversity inside the population^[@6920034, p.1]
* Higher chance for "specialist" solutions to be selected since it is decided on a per case basis
* Performance increases have been demonstrated in many benchmarking problems ^[@epsilon_lexicase_main, p.744-745]

@@

### Hypothesis testing

#### TODO: reformulate hypothesis
1. The usage of $\epsilon$-lexicase selection will result in models that perform significantly different than models that are evolved using tournament selection:
2. Statistical significant differences between the mean errors of training and testing data exist for both selection operators
3. No differences in program size exist between the distribution underlying the samples produced by $\epsilon$-lexicase and the distribution underlying samples of tournament selection

### 2022-06-26 docs/rmd/presentation/slides.tex
...

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  ignorenonframetext,
]{beamer}
\usepackage{pgfpages}
\setbeamertemplate{caption}[numbered]
\setbeamertemplate{caption label separator}{: }
\setbeamercolor{caption name}{fg=normal text.fg}
\beamertemplatenavigationsymbolsempty
% Prevent slide breaks in the middle of a paragraph
\widowpenalties 1 10000
\raggedbottom
\setbeamertemplate{part page}{
  \centering
  \begin{beamercolorbox}[sep=16pt,center]{part title}
    \usebeamerfont{part title}\insertpart\par
  \end{beamercolorbox}
}
\setbeamertemplate{section page}{
  \centering
  \begin{beamercolorbox}[sep=12pt,center]{part title}
    \usebeamerfont{section title}\insertsection\par
  \end{beamercolorbox}
}
\setbeamertemplate{subsection page}{
  \centering
  \begin{beamercolorbox}[sep=8pt,center]{part title}
    \usebeamerfont{subsection title}\insertsubsection\par
  \end{beamercolorbox}
}
\AtBeginPart{
  \frame{\partpage}
}
\AtBeginSection{
  \ifbibliography
  \else
    \frame{\sectionpage}
  \fi
}
\AtBeginSubsection{
  \frame{\subsectionpage}
}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usecolortheme{beaver}
\usefonttheme{structurebold}
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\newif\ifbibliography
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {}%
  {\par}

\title{Analyzing the influence of Selection on Genetic Programming's
Generalization ability in Symbolic Regression}
\subtitle{A comparison of epsilon-lexicase Selection and Tournament
Selection}
\author{}
\date{\vspace{-2.5em}2022-06-29}

@@

\hypertarget{research-question}{%
\section{Research Question}\label{research-question}}

\begin{frame}{Research Question}
\begin{itemize}
\tightlist
\item
  Does the usage of \(\epsilon\)-lexicase parent selection influence the
  generalization behaviour of genetic programming in symbolic regression
  if compared to tournament selection?
\item
\end{itemize}
\end{frame}

@@

\begin{frame}{Parent Selection}
\protect\hypertarget{parent-selection}{}
\begin{itemize}
\tightlist
\item
  Operator that selects individual solutions from the population for
  reproduction and mutation
\item
  Most commonly used selection operator in GP: Tournament
  selection\footnote<.->{Fang and Li (2010), p.181}
\item
  Intuition: High chance for ``generalist'' solutions to be selected
  since it is based on aggregated fitness scores
\end{itemize}
\end{frame}

@@

\begin{frame}{Symbolic Regression}
\protect\hypertarget{symbolic-regression}{}
\begin{itemize}
\tightlist
\item
  Task: Find a mathematical model that fits a given set of datapoints
\item
  One of the first applications of Genetic Programming introduced
\end{itemize}

\begin{frame}{Generalization}
\protect\hypertarget{generalization}{}
\end{frame}

\begin{frame}{Motivation}
\protect\hypertarget{motivation}{}
\begin{itemize}
\tightlist
\item
  Generalization
\end{itemize}
\end{frame}

\begin{frame}{Genetic Programming}
\protect\hypertarget{genetic-programming-1}{}
\ldots{}

@@

\begin{frame}{Experimental Study}
\ldots{}
\end{frame}

\begin{frame}{Research Design}
\protect\hypertarget{research-design}{}
\ldots{}
\end{frame}

@@

@@

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-10.1007ux2f978-3-642-16493-4_19}{}%
Fang, Y. and Li, J. (2010) `A review of tournament selection in genetic
programming', in Cai, Z. et al. (eds) \emph{Advances in computation and
intelligence}. Berlin, Heidelberg: Springer Berlin Heidelberg, pp.
181--192.

@@

\leavevmode\hypertarget{ref-epsilon_lexicase_main}{}%
La Cava, W., Spector, L. and Danai, K. (2016) `Epsilon-lexicase
selection for regression', in \emph{Proceedings of the genetic and
evolutionary computation conference 2016}. New York, NY, USA:
Association for Computing Machinery (GECCO '16), pp. 741--748.
doi:\href{https://doi.org/10.1145/2908812.2908898}{10.1145/2908812.2908898}.
\end{cslreferences}
\end{frame}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  ignorenonframetext,
]{beamer}
\usepackage{pgfpages}
\setbeamertemplate{caption}[numbered]
\setbeamertemplate{caption label separator}{: }
\setbeamercolor{caption name}{fg=normal text.fg}
\beamertemplatenavigationsymbolsempty
% Prevent slide breaks in the middle of a paragraph
\widowpenalties 1 10000
\raggedbottom
\setbeamertemplate{part page}{
  \centering
  \begin{beamercolorbox}[sep=16pt,center]{part title}
    \usebeamerfont{part title}\insertpart\par
  \end{beamercolorbox}
}
\setbeamertemplate{section page}{
  \centering
  \begin{beamercolorbox}[sep=12pt,center]{part title}
    \usebeamerfont{section title}\insertsection\par
  \end{beamercolorbox}
}
\setbeamertemplate{subsection page}{
  \centering
  \begin{beamercolorbox}[sep=8pt,center]{part title}
    \usebeamerfont{subsection title}\insertsubsection\par
  \end{beamercolorbox}
}
\AtBeginPart{
  \frame{\partpage}
}
\AtBeginSection{
  \ifbibliography
  \else
    \frame{\sectionpage}
  \fi
}
\AtBeginSubsection{
  \frame{\subsectionpage}
}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usecolortheme{beaver}
\usefonttheme{structurebold}
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression},
  pdfauthor={Roman Hoehn, B.Sc. Wirtschaftspaedagogik},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\newif\ifbibliography
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {}%
  {\par}

\title{Analyzing the influence of Selection on Genetic Programming's
Generalization ability in Symbolic Regression}
\subtitle{A comparison of epsilon-lexicase Selection and Tournament
Selection}
\author{Roman Hoehn, B.Sc. Wirtschaftspaedagogik}
\date{2022-06-29}

@@

\begin{frame}{Research Question}
\protect\hypertarget{research-question}{}
\begin{itemize}
\tightlist
\item
  Does the usage of \(\epsilon\)-lexicase parent selection influence the
  generalization behaviour of genetic programming in symbolic regression
  if compared to tournament selection?
\end{itemize}
\end{frame}

@@

\begin{frame}{Parent Selection}
\protect\hypertarget{parent-selection}{}
\begin{itemize}
\tightlist
\item
  Operator that selects individual solutions from the population for
  reproduction and mutation
\item
  Most commonly used selection operator in Genetic Programming (GP):
  Tournament selection\footnote<.->{Fang and Li (2010), p.181}
\item
  Intuition: High chance for ``generalist'' solutions to be selected
  since it is based on aggregated fitness scores
\end{itemize}
\end{frame}

@@

\begin{frame}{Symbolic Regression}
\protect\hypertarget{symbolic-regression}{}
\begin{itemize}
\tightlist
\item
  Task: Find a mathematical model that fits a given set of datapoints
\item
  One of the first applications of GP described by Koza (1992)
\item
  High relevance: GP can outperform state-of-the-art machine learning
  algorithms like gradient boosting\footnote<.->{Orzechowski, Cava and
    Moore (2018)}
\end{itemize}
\end{frame}

\begin{frame}{Generalization}
\protect\hypertarget{generalization}{}
\begin{itemize}
\tightlist
\item
  The ability of a model to perform well on previously unseen fitness
  cases
\item
  Main objective in most supervised machine learning problems
\item
  Challenge: Avoid overfitting to training data
\end{itemize}
\end{frame}

\begin{frame}{Motivation}
\protect\hypertarget{motivation}{}
\begin{itemize}
\tightlist
\item
  Little attention has been paid to generalization in GP\footnote<.->{O'Neill
    \emph{et al.} (2010), Kushchu (2002)}
\item
  High practical relevance of symbolic regression in many fields,
  e.g.~financial forecasting
\end{itemize}
\end{frame}

@@

\begin{frame}{Benchmark problem}
\protect\hypertarget{benchmark-problem}{}
UC Irvine Machine Learning Repository: Prediction of energy efficiency
in buildings\footnote<.->{Dua and Graff (2017)}

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-1}Overview - Energy Heating data set}
\centering
\begin{tabular}[t]{l|l}
\hline
\textbf{Variable} & \textbf{Description}\\
\hline
X1 & Relative Compactness\\
\hline
X2 & Surface Area\\
\hline
X3 & Wall Area\\
\hline
X4 & Roof Area\\
\hline
X5 & Overall Height\\
\hline
X6 & Orientation\\
\hline
X7 & Glazing Area\\
\hline
X8 & Glazing Area Distribution\\
\hline
y1 & Heating Load\\
\hline
y2 & Cooling Load\\
\hline
\end{tabular}
\end{table}
\end{frame}

\begin{frame}{Experiment}
\protect\hypertarget{experiment}{}
\begin{block}{Single run}
\protect\hypertarget{single-run}{}
\begin{itemize}
\tightlist
\item
  Total dataset (\(N=768\)) is randomly split into a training and
  testing dataset (50:50)
\item
  Train two models using GP with the training dataset only, one using
  tournament selection and the other \(\epsilon\)-lexicase selection
\item
  For each generation: Select elite model and compute its fitness for
  the testing dataset
\end{itemize}
\end{block}

\begin{block}{Full experiment}
\protect\hypertarget{full-experiment}{}
\begin{itemize}
\tightlist
\item
  Stochastic algorithm: Repeat the basic experiment for 50 total runs
\item
  Collect and aggregate results for training error, testing error and
  program length
\end{itemize}
\end{block}
\end{frame}

@@

\begin{frame}{Research Question}
\protect\hypertarget{research-question-1}{}
\begin{itemize}
\tightlist
\item
  Does the usage of \(\epsilon\)-lexicase parent selection influence the
  generalization behaviour of genetic programming in symbolic regression
  if compared to tournament selection?
\end{itemize}

\begin{block}{Hypothesis testing}
\protect\hypertarget{hypothesis-testing}{}
\begin{block}{TODO: reformulate hypothesis}
\protect\hypertarget{todo-reformulate-hypothesis}{}
\begin{enumerate}
\tightlist
\item
  The usage of \(\epsilon\)-lexicase selection will result in models
  that perform significantly different than models that are evolved
  using tournament selection:
\item
  Statistical significant differences between the mean errors of
  training and testing data exist for both selection operators
\item
  No differences in program size exist between the distribution
  underlying the samples produced by \(\epsilon\)-lexicase and the
  distribution underlying samples of tournament selection
\end{enumerate}
\end{block}
\end{block}
\end{frame}

@@

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-Dua:2019}{}%
Dua, D. and Graff, C. (2017) `UCI machine learning repository'.
University of California, Irvine, School of Information; Computer
Sciences. Available at: \url{http://archive.ics.uci.edu/ml}.

\leavevmode\hypertarget{ref-10.1007ux2f978-3-642-16493-4_19}{}%
Fang, Y. and Li, J. (2010) `A review of tournament selection in genetic
programming', in Cai, Z. et al. (eds) \emph{Advances in computation and
intelligence}. Berlin, Heidelberg: Springer Berlin Heidelberg, pp.
181--192.

@@

\leavevmode\hypertarget{ref-generalisation_in_gp}{}%
Kushchu, I. (2002) `An evaluation of evolutionarygeneralisation in
genetic programming', \emph{Artificial Intelligence Review - AIR}, 18,
pp. 3--14.
doi:\href{https://doi.org/10.1023/A:1016379201230}{10.1023/A:1016379201230}.

\leavevmode\hypertarget{ref-epsilon_lexicase_main}{}%
La Cava, W., Spector, L. and Danai, K. (2016) `Epsilon-lexicase
selection for regression', in \emph{Proceedings of the genetic and
evolutionary computation conference 2016}. New York, NY, USA:
Association for Computing Machinery (GECCO '16), pp. 741--748.
doi:\href{https://doi.org/10.1145/2908812.2908898}{10.1145/2908812.2908898}.

\leavevmode\hypertarget{ref-open_issues_gp}{}%
O'Neill, M. \emph{et al.} (2010) `Open issues in genetic programming',
\emph{Genetic Programming and Evolvable Machines}, 11, pp. 339--363.
doi:\href{https://doi.org/10.1007/s10710-010-9113-2}{10.1007/s10710-010-9113-2}.

\leavevmode\hypertarget{ref-Orzechowski_2018}{}%
Orzechowski, P., Cava, W.L. and Moore, J.H. (2018) `Where are we now?',
in \emph{Proceedings of the genetic and evolutionary computation
conference}. ACM.
doi:\href{https://doi.org/10.1145/3205455.3205539}{10.1145/3205455.3205539}.
\end{cslreferences}
\end{frame}
